---
title: "Case study on Covid-19 B-cells: differential detection analysis"
author: "Jeroen Gilis & Laura Perin"
date: "14/11/2023"
output:
    html_document:
      code_download: true    
      theme: cosmo
      toc: true
      toc_float: true
      highlight: tango
      number_sections: true
---

# Load libraries

```{r, message=FALSE, warning=FALSE}
library(SingleCellExperiment)
library(scuttle)
library(ggplot2)
library(patchwork)
library(scater)
library(scran)
library(edgeR)
library(limma)
library(pbapply)
```

```{r}
here::i_am("3_CovidCase_DD.Rmd")
```

# Load DGE analysis data and results

```{r}
pb_ct <- readRDS("./objects/pb_ct_filt.rds")
DGE_res <- readRDS("./objects/DGE_res_edgeR_NB.rds")
```

# Load sce data

```{r}
sce <- readRDS("./objects/sce_Covid_Bcells.rds")
```

# Binarize and pseudobulk

```{r}
# binarize
assays(sce)$detection <- assays(sce)$counts
assays(sce)$detection[assays(sce)$detection >=1] <- 1

# aggregate
pb_bin <- aggregateAcrossCells(sce,
                               use.assay.type = "detection",
                               ids = colData(sce)[,c("donor_id","cell_type_curated")])
pb_bin$Status_on_day_collection_summary <- as.factor(pb_bin$Status_on_day_collection_summary)

pb_bin$cell_type_curated <- as.factor(pb_bin$cell_type_curated)
```

# Filtering

Retain the same five cell types

```{r}
pb_bin <- pb_bin[,pb_bin$cell_type_curated %in% c("B cell", "class switched memory B cell",
                                      "immature B cell", "naive B cell",
                                      "unswitched memory B cell")]
colData(pb_bin) <- droplevels(colData(pb_bin))
```

Add a column for celltype-donor combination -> unique sample identifier

```{r}
for (i in seq_along(pb_ct)) {
  pb_ct[[i]]$sample_id <- paste0(pb_ct[[i]]$donor_id, 
                                 "_", 
                                 gsub(" ", "_", pb_ct[[i]]$cell_type_curated))
}
pb_bin$sample_id <- paste0(pb_bin$donor_id, 
                           "_", 
                           gsub(" ", "_", pb_bin$cell_type_curated))
```

Copy the gene-level and patient-level filtering from the DGE analysis.

```{r}
label <- pb_bin$cell_type_curated
pb_bin_ct <- vector(mode =  "list", length = length(unique(label)))

for (i in 1:length(unique(label))) {
  
  # currentlab <- as.character(sort(unique(label))[i])
  # # added line -> essential, as sample_ is only unique within cell type
  # pb_bin_hlp <- pb_bin[,pb_bin$cell_type_curated == currentlab]
  
  pb_bin_ct[[i]] <- pb_bin[rownames(pb_ct[[i]]),
                           match(pb_ct[[i]]$sample_id, pb_bin$sample_id)]
  colData(pb_bin_ct[[i]]) <- droplevels(colData(pb_bin_ct[[i]]))
}
names(pb_bin_ct) <- sort(unique(label))
```

```{r}
saveRDS(pb_bin_ct, "./objects/pb_bin_ct_filt.rds")
```

# Setup contrast matrix

```{r}
design <- model.matrix(~ Site + sex + Status_on_day_collection_summary, 
                         colData(pb_bin_ct$`B cell`))
colnames(design) <- sub("Status_on_day_collection_summary", "", colnames(design))
colnames(design)[1] <- "Intercept"
  
L <- matrix(0, ncol = 5, nrow = ncol(design)) # initialize contrast matrix
rownames(L) <- colnames(design)
colnames(L) <- c("AsymptomaticVsHealthy", "MildVsHealthy", 
                 "ModerateVsHealthy", "SevereVsHealthy",
                 "CriticalVsHealthy")

L["Asymptomatic",1] <- 1
L["Mild",2] <- 1
L["Moderate",3] <- 1
L["Severe",4] <- 1
L["Critical",5] <- 1
L
```

# Differential detection analysis - edgeR-NB_optim

## Perform analysis

```{r}
DD_res <- pblapply(pb_bin_ct, function(element){
  
    # get cell type name
    ct <- levels(droplevels(element$cell_type_curated))
  
    # create design matrix
    design <- model.matrix(~ Site + sex + Status_on_day_collection_summary, 
                           colData(element))
    colnames(design) <- sub("Status_on_day_collection_summary", "", colnames(design))
    colnames(design)[1] <- "Intercept"
    
    # Remove features detected in (nearly) all cells
    med_detection <- rowMedians(sweep(assay(element), 2, element$ncells, "/"))
    hist(med_detection, breaks=40)
    abline(v=0.9, col="red", lwd=2, cex.main=0.7)
    
    # Differential detection analysis with edgeR-NB_optim
    y <- DGEList(counts=assay(element)[med_detection < 0.9,], samples=colData(element))
    of <- colMeans(sweep(assay(element)[med_detection < 0.9,], 2, element$ncells, "/"))
    y$offset <- log(element$ncells*of)
    y <- estimateDisp(y, design)
    fit <- glmQLFit(y, design, robust=T)
  
    # Intermediate visualization
    par(mfrow=c(1,2))
    plotBCV(y, main = paste0("QLDisp plot for ", ct), cex.main=0.7)
    plotQLDisp(fit, main = paste0("QLDisp plot for ", ct), cex.main=0.7)
    abline(h=1, col="gold4", lwd=2)
    
    # Differential testing
    resList_i <- vector(mode = "list", length = ncol(L))
    for(j in 1:ncol(L)){
      test <- glmQLFTest(fit, contrast=L[,j], poisson.bound = FALSE)
      resList_i[[j]] <- test$table
    }
    return(resList_i)
})
```

Note: by removing genes that are detected in (nearly) all samples in the
differential detection analysis but not the differential expression analysis,
the results will be of different dimensions. For easy comparison between the two
analyses, I will here fill in `NA` values for the genes that were removed in the
DD analysis

```{r}
for(i in seq_along(DD_res)){
  for(j in 1:5){
    removed <- which(!rownames(DGE_res[[i]][[j]]) %in% rownames(DD_res[[i]][[j]])) # detect removed genes
    DD_res[[i]][[j]][nrow(DD_res[[i]][[j]]) + length(removed),] <- NA # add NA values for removed genes
    rownames(DD_res[[i]][[j]])[(nrow(DD_res[[i]][[j]])-length(removed)+1):nrow(DD_res[[i]][[j]])] <- rownames(DGE_res[[i]][[j]])[removed] # add rownames for removed genes
    DD_res[[i]][[j]] <- DD_res[[i]][[j]][match(rownames(DGE_res[[i]][[j]]), rownames(DD_res[[i]][[j]])),] # correct order    
  }

}

saveRDS(DD_res, "./objects/DD_res_edgeR_NB_optim.rds")
```

## Visualize p-values

```{r}
pdf("./figures/pval_hist_DD_edgeR_NB_optim.pdf",
            width     = 8,
            height    = 8,
            pointsize = 4)
par(mfrow=c(5,5))
for (j in 1:5) {
  for(i in 1:5){
    pat_per_ct <- table(pb_bin_ct[[j]]$Status_on_day_collection_summary)
    hist(DD_res[[j]][[i]]$PValue, breaks=20, xlab = "p-values",
         main = paste0(levels(droplevels(pb_bin_ct[[j]]$cell_type_curated)), " \n ", 
                       levels(droplevels(pb_bin_ct[[j]]$Status_on_day_collection_summary))[i+1], " - ",
                       pat_per_ct[1], " vs ",pat_per_ct[i+1]))
  }
}
dev.off()
```

# Session info

```{r}
sessionInfo()
```
